---
title: "Project Summary"
subtitle: "Project name"
author: "Student 1, Student 2, ..."
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#read the data from github and pull in libraries

library(ggplot2)
library(readr)

d2p1 <- read_csv(url("https://raw.githubusercontent.com/chrisgraham631/D2P/main/d2p1.csv"))
food <- read_csv(url("https://raw.githubusercontent.com/chrisgraham631/D2P/main/foodaccess3.csv"))
```

# Introduction

Relevance, question of interest, background, etc.

# Data cleaning

The data cleaning process never really ended.  Initially everyone collected relevant data and a large data table was created with most of the variables we used.  As the analysis continued, we had to add a few additional variables.  The main dataset used broke down the data by state, but additional data was analyzed at the both the national level and the Colorado county level.  

The initial data cleaning required using many techniques learned in the Module 4 and a few that were found on the internet.  Common functions like select, filter and rename were used frequently.  The mutate function was used multiple times.  Mostly it was used to create either a proportion of a variable from a total population or to create a population estimate based on a proportion.  One of the datasets had the individual states broken down by something called a CoC number.  This required us to the separate function to separate the state abbreviation from the CoC number.  This left us with multiple rows for each state and a few extra rows with numeric data.  After doing some digging, we found the aggregate function, which allowed us to combine the various rows for each state by summing the numeric data.  Once the various data tables were clean and named correctly, the merge function was used to combine the data into one table.  It was then written as a .csv.

# Numeric summaries

Provide a numeric summary of the response and at least 3 other variables along with a brief interpretation of each summary.

* summarize Latino and Black (Chris)

# Univariate graphics

Provide graphical summaries of the response and at least 3 other variables along with a brief interpretation of each graphic.

Response:
```{r}
ggplot(d2p1, aes(x=youth_pov_prop_est)) + geom_histogram()
```

* histogram of youth poverty and show where colorado falls (Chris)

Predictors:
```{r}
#I wanted the violin plot of the entire United States but this plot was requiring a y aesthetic.  CPI is the same for all 50 states so it was chosen for the y aesthetic.
ggplot(d2p1) + geom_violin(aes(x = prop_latino_youth_pov, y = cpi)) + geom_boxplot(aes(x = prop_latino_youth_pov, y = cpi), width = 0.1) + xlab("x") + ylab("y") + labs(title = "Proportion of Latino youth living in poverty")
```
The graphic above has a violin plot with a small boxplot on top.  This gives us information about the distribution of the data.  The boxplot shows the Q1, median and Q3 values, while the violin plot shows the density of the data.  The violin plot would also show if the data is multimodal.  It appears as if the mode is around 0.24 but a small second mode may be around 0.34.  From this we can see that 50% of states have less than 25.5% of their Latino youth population in poverty and 75% of the states have less than 33% of their Latino youth population living in poverty.  It also shows a maximum value of 40% of Latino youth living in poverty and a minimum value of around 11%.


```{r}
ggplot(d2p1) + geom_violin(aes(x = prop_black_youth_pov, y = cpi)) + geom_boxplot(aes(x = prop_black_youth_pov, y = cpi), width = 0.1) + xlab("x") + ylab("y") + labs(title = "Proportion of Black youth living in poverty")
```
The graphic above has a violin plot with a small boxplot on top.  This gives us information about the distribution of the data.  The boxplot shows the Q1, median and Q3 values, while the violin plot shows the density of the data.  The violin plot would also show if the data is multimodal.  The mode of this dataset appears to be around 0.38 with a potential smaller second mode around 0.27.  The boxplot shows that 25% of states have less than 27% of their black youth population living in poverty while 50% of states have less than 35% of their black youth living in poverty.  For this plot the maximum percentage of black youth living in poverty is around 47% and the minimum is around 17%.  The state with the lowest proportion of black youth living in poverty is still at nearly 20%, thats way too high.
* box plot health insurance
* violin of:
  * latino
  * 

# Bivariate graphics

Provide bivariate graphical summaries of the response versus at least 3 other variables along with a brief interpretation of each graphic.

* state map showing youth poverty rates (chris)
* youth poverty vs food access scatter plot
* health insurance rates on us map (chris)

# Multivariate graphics

Provide at least 3 multivariate graphical summaries along with a brief interpretation of each graphic. This could be something like a scatter plot of the response versus another variable that facets by a third variable (or uses colors or shapes to distinguish levels of another variable).

* bar chart showing ethnicity, total, and poverty.
* scatter plot showing health insurance vs. youth poverty vs. states
* **find another** (chris)

# Conclusions and recommendations

What are your overall conclusions about the relationship between your response variable and your other variables? Based on this information, what policy recommendations would you propose to address your findings?

In conclusion, although Colorado ranks well in statewide youth poverty rate compared to the rest of the Unite States, one in ten children still fall below the poverty line. Yet, it has long been a challenge to determine what factors are both influential and targetable for some type of policy change in regards to youth poverty rates. Our model contained three powerful predictors for youth poverty rate. The percentage of a State's population that is covered by some form of health insurance explains more of the variance in a State's youth poverty rate than the other two predictors, as such it is the best key to bringing down youth poverty. More precisely, holding the percentage of Latino and Black youth in poverty constant, a 1% increase in health insurance coverage returns a 0.5% decrease in youth poverty. This implies that if 100% of the population of Colorado is covered with health insurance, our model estimates that the State’s youth poverty rate will decline from 11% to 5%. 

# Appendix

The code you used to clean and analyze your data
