---
title: "Project Summary"
subtitle: "Holistic Approach to Understanding and Combating Child Poverty in Colorado"
author: "Bai Jaitrong, Christopher Graham, David Palcsak"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
#read the data from github and pull in libraries

library(ggplot2)
library(readr)

d2p1 <- read_csv(url("https://raw.githubusercontent.com/chrisgraham631/D2P/main/d2p1.csv"))
food <- read_csv(url("https://raw.githubusercontent.com/chrisgraham631/D2P/main/foodaccess3.csv"))
eth <- read_csv(url("https://raw.githubusercontent.com/chrisgraham631/D2P/main/national_ethnicity_histogram_data.csv"))

#data manipulation used by Bai, changing crime variable to crime rate per 100K

d2pbai <- d2p1 %>% mutate(prop_health_insurance = health_insurance/total_pop*100, prop_educ_bach = educ_bach/total_pop*100, violent_crime_per_100K = violent_crime/population *100000, homicide_per_100K = homicide/population*100000, raped_revised_per_100K = rape_revised/population*100000, robbery_per_100K = robbery/population *100000, aggravated_assault_per_100K = aggravated_assault/population *100000, property_crime_per_100K = property_crime/population *100000, burgulary_per_100K = burglary/population *100000, larceny_per_100K = larceny/population*100000, motor_vehicle_theft_per_100K = motor_vehicle_theft/population *100000) %>% select(State_abb, youth_pov_prop_est,unemployment, prop_health_insurance,prop_educ_bach,elem_size,second_size,salary, prop_latino_youth_pov,prop_asian_youth_pov,prop_black_youth_pov, prop_american_indian_youth_pov, prop_white_youth_pov, prop_other_youth_pov,violent_crime_per_100K,homicide_per_100K,raped_revised_per_100K, robbery_per_100K,aggravated_assault_per_100K,property_crime_per_100K,burgulary_per_100K,larceny_per_100K,motor_vehicle_theft_per_100K)

```

# Introduction

In 2017 the United States ranked 6th in GDP per capita among the 36 countries in The Organization for Economic Co-operation and Development (OECD).  In that same year the United States had the third highest proportion of youth in poverty. Children raised in poverty are more likely to live in poverty in their adult lives, and raise any children they have in poverty, thus continuing the “Cycle of Poverty”.

Our objective was to explore what socioeconomic factors have an effect on childhood poverty in the United States and suggest policy changes that can be made in Colorado to help reduce our state's poverty rate and break the cycle?

We collected data from various federal, state, and non-profit agencies such as HUD and the Annie E. Casey Foundation’s website *The Kids Count Data Center*. Our final data set consisted of 51 observations and 50 variables related to poverty, unemployment, health insurance coverage, education, crime rate, food access, and ethnicity for each of the states and DC.

Using exploratory data analysis, we determined which variables had a linear relationship with youth poverty. We further narrowed down the variables by using linear regression analysis to determine which of the 50 variables were the most influential in predicting youth poverty rates in our multiple regression model. 

We tested our model assumptions of homoscedasticity, linearity, normality, no multicollinearity, and no autocorrelation using a combination of formal testing and visual inspection of scatter plots. For example, We use the Durbin-Watson test to confirm that there were no autocorrelation in the model. We removed any variables with a variance inflation factor (VIF) greater than 5 to minimize any multicollinearity in the model. We also remove any variables with a p-value great greater than.05. This produced a model with 3 explanatory variables, prop_health_insurance, percent_latino_youth_pov, and percent_black_youth_pov.

# Data cleaning

The data cleaning process never really ended.  Initially everyone collected relevant data and a large data table was created with most of the variables we used.  As the analysis continued, we had to add a few additional variables.  The main dataset used broke down the data by state, but additional data was analyzed at the both the national level and the Colorado county level.  

The initial data cleaning required using many techniques learned in the Module 4 and a few that were found on the internet.  Common functions like select, filter and rename were used frequently.  The mutate function was used multiple times.  Mostly it was used to create either a proportion of a variable from a total population or to create a population estimate based on a proportion.  One of the datasets had the individual states broken down by something called a CoC number.  This required us to the separate function to separate the state abbreviation from the CoC number.  This left us with multiple rows for each state and a few extra rows with numeric data.  After doing some digging, we found the aggregate function, which allowed us to combine the various rows for each state by summing the numeric data.  Once the various data tables were clean and named correctly, the merge function was used to combine the data into one table.  It was then written as a .csv.

# Numeric summaries

Provide a numeric summary of the response and at least 3 other variables along with a brief interpretation of each summary.

#Youth Poverty Rate 

```{r}

summary(d2p1$youth_pov_prop_est)
sd(d2p1$youth_pov_prop_est)
sd(d2p1$youth_pov_prop_est)/15.93

```

The median and mean of the youth poverty  rate are 15.50% and 15.93% respectively. The mean is larger than the median indicating that the data is skewed to the right. The lowest youth poverty rate is at 8.10% and the highest poverty rate is at 27.60 %. Based on the ratio of the standard deviation and mean being approximately 27%, the spread of the data would be considered low.

#Proportion of the population who has Health Insurance Coverage 
```{r}

summary(d2p1$prop_health_insurance)
sd(d2p1$prop_health_insurance)
sd(d2p1$prop_health_insurance)/89.17

```

The median and mean of the proportion of the population who has health insurance coverage is 89.45% and 89.17%. The mean is smaller than the median indicting that the data is skewed to the left. The lowest proportion of the population who has health insurance coverage is 80.75% and the highest is 96.75%. Based on the ratio of the standard deviation and mean being approximately 3.8%, the spread of the data would be considered low. 

* summarize Latino and Black (Chris)

# Univariate graphics

### Response Variable:
```{r message=FALSE}
ggplot(d2p1, aes(x=youth_pov_prop_est)) + geom_histogram() + geom_vline(aes(xintercept=d2p1$youth_pov_prop_est[d2p1$State_abb=='CO'], color="red", linetype = "dashed")) +
geom_text(aes(x=d2p1$youth_pov_prop_est[d2p1$State_abb=='CO'], y=4.5, label = "Colorado\n", angle = 90, color = "red")) +
theme_bw() +
theme(legend.position = "none") +
xlab("Youth Poverty Rate ") + 
ylab("Count")
```

Colorado had a youth poverty rate of 11.2% in our dataset.  From the graphic above we can conclude that Colorado is considerably low in its youth poverty rate compared to other states.  In fact, Colorado is in the 6th percentile and 6th overall in the United States. 

### Predictor Variables:
```{r}
#I wanted the violin plot of the entire United States but this plot was requiring a y aesthetic.  CPI is the same for all 50 states so it was chosen for the y aesthetic.
ggplot(d2p1) + geom_violin(aes(x = prop_latino_youth_pov, y = cpi)) + geom_boxplot(aes(x = prop_latino_youth_pov, y = cpi), width = 0.1) + xlab("x") + ylab("y") + labs(title = "Proportion of Latino youth living in poverty") +
theme_bw()
```

The graphic above has a violin plot with a small boxplot on top.  This gives us information about the distribution of the data.  The boxplot shows the Q1, median and Q3 values, while the violin plot shows the density of the data.  The violin plot would also show if the data is multimodal.  It appears as if the mode is around 0.24 but a small second mode may be around 0.34.  From this we can see that 50% of states have less than 25.5% of their Latino youth population in poverty and 75% of the states have less than 33% of their Latino youth population living in poverty.  It also shows a maximum value of 40% of Latino youth living in poverty and a minimum value of around 11%.


```{r}
ggplot(d2p1) + geom_violin(aes(x = prop_black_youth_pov, y = cpi)) + geom_boxplot(aes(x = prop_black_youth_pov, y = cpi), width = 0.1) + xlab("x") + ylab("y") + labs(title = "Proportion of Black youth living in poverty") +
theme_bw() 
```

The graphic above has a violin plot with a small boxplot on top.  This gives us information about the distribution of the data.  The boxplot shows the Q1, median and Q3 values, while the violin plot shows the density of the data.  The violin plot would also show if the data is multimodal.  The mode of this dataset appears to be around 0.38 with a potential smaller second mode around 0.27.  The boxplot shows that 25% of states have less than 27% of their black youth population living in poverty while 50% of states have less than 35% of their black youth living in poverty.  For this plot the maximum percentage of black youth living in poverty is around 47% and the minimum is around 17%.  The state with the lowest proportion of black youth living in poverty is still at nearly 20%, thats way too high.

* box plot health insurance


# Bivariate graphics

```{r}
#proportion of youth in poverty vs crime rate
ggplot(d2p1, aes(y = youth_pov_prop_est, x = violent_crime_per_100K)) + 
  geom_point() +
  geom_smooth(method = lm, se = FALSE)+
  labs(title = "Percentage of youth in poverty vs Violent Crime Rate") +
  ylab("Percentage of Youth in Poverty") + 
  xlab("Violent Crimes per 100,000 People") +
theme_bw() 
```
This is a scatter plot of the percentage of youth in poverty vs violent crimes per 100,000 people.  The trend line shows a positive linear relationship which indicates that as one rate rises, so does the other.  The slope of the line is less than one indicating the violent crime rate increases at a faster rate than the percentage of youth in poverty does.  The data points fall near the line so the relationship likely exists, however, there is still a fair amount of variance indicating this may not be the strongest relationship.

*Provide bivariate graphical summaries of the response versus at least 3 other variables along with a brief interpretation of each graphic.*

* state map showing youth poverty rates (chris)

* youth poverty vs food access scatter plot (Bai)

```{r, echo=FALSE}

food <-  food %>% %>% mutate(perc_weighted_mean_lakids1share = weighted_mean_lakids1share*100, youth_pov_rate = youth_poverty_rate_2010*100) 

ggplot(food,aes(x = perc_weighted_mean_lakids1share, y = youth_pov_rate)) + 
  geom_smooth(method = lm,se = FALSE) +
  xlab("Percentage of youth living in low food access areas") + 
  ylab("2010 Youth Poverty Rate")


```

The scatter plot of the 2010 youth poverty rate vs percentage of youth living in low food access areas shows that there is a strong positive relationship between the two variables. As the percentage of youth living in low food access areas increases, youth poverty rate increases. 

# Multivariate graphics

Provide at least 3 multivariate graphical summaries along with a brief interpretation of each graphic. This could be something like a scatter plot of the response versus another variable that facets by a third variable (or uses colors or shapes to distinguish levels of another variable).

* bar chart showing ethnicity, total, and poverty.


# scatter plot showing health insurance vs. youth poverty vs. states

```{r, warning= False, comment= " "}

ggplot(d2p1,aes(prop_health_insurance,youth_pov_prop_est)) + geom_point( aes(col=State_abb)) + geom_smooth(method = lm, se= FALSE) + geom_text(aes(label = State_abb)) + theme(legend.position = "none")

```

The above scatter plot indicates that the youth poverty rate decreases as the proportion of the population who has health insurance coverage increases. It also shows that the colder climate states have a higher proportion of the population with health insurance coverage and with a lower youth poverty rate. 


* **find another** (chris)

```{r, fig.width = 12, fig.height = 2}
ggplot(eth, aes(fill = Ethnicity, x = Type, y = Proportion)) + 
  geom_bar(position = "stack", stat = "identity") +
  labs(title = "Youth Population and Youth Population in Poverty by Ethnicity") +
  coord_flip() +
theme_bw() 
```
The graphic above shows the ethnic breakdown of the total population of children in the United States of America and the ethnic breakdown of children who live in poverty in the USA.  If ethnicity had no effect on children living in poverty, then we would expect each ethnic group to have the same representation in both bars.  While the graphic does not allow for quantitative analysis, it does show whether youth in certain ethnic groups are more or less likely to be exposed to poverty.  The bar representing the 'Non-Hispanic White' group represents a significantly bigger portion of the total childhood population than it does the population of children in poverty which indicates that children in the 'Non-Hispanic White' ethnic group are less likely to be exposed to poverty.  Children in the 'Asian and Pacific Islander' group are less likely to be exposed to poverty by the same observation.  This graphic also shows that children in the 'Hispanic or Latino', 'Black or African American' and 'American Indian' ethnic groups are more likely to be exposed to poverty because they represent a larger portion of the population of children in poverty than the total population of children.

# Conclusions and recommendations

In conclusion, although Colorado ranks well in statewide youth poverty rate compared to the rest of the Unite States, one in ten children still fall below the poverty line. Yet, it has long been a challenge to determine what factors are both influential and targetable for some type of policy change in regards to youth poverty rates. Our model contained three powerful predictors for youth poverty rate. The percentage of a State's population that is covered by some form of health insurance explains more of the variance in a State's youth poverty rate than the other two predictors, as such it is the best key to bringing down youth poverty. More precisely, holding the percentage of Latino and Black youth in poverty constant, a 1% increase in health insurance coverage returns a 0.5% decrease in youth poverty. This implies that if 100% of the population of Colorado is covered with health insurance, our model estimates that the State’s youth poverty rate will decline from 11% to 5%. 

Based on our analysis, our policy recommendation is two-fold. First, Government funded assistance programs play a major role in reducing poverty, specifically for Black and Latino children. Studies have shown that increasing Earned Income Tax credits by 40% can decrease youth poverty by 16% nation-wide. The Supplemental Nutrition Assistance Program was expanded in 2010 and one study showed a decrease in youth poverty of 16.2%. Black and Hispanic workers are about twice as likely to work for poverty wages so raising the minimum wage would help target the highest risk minority groups according to our model. In general, the more funding we can give government assistance programs, the less youth we would expect to see in poverty. Our second recommendation is to pass some form of universal health care. According to our model, this is the best thing we can do to reduce childhood poverty. Ideally, we would pass this on a national level. A recent survey from the Pew Research Center showed that 62% of Americans now favor a single payer health care system. The biggest public barrier is agreeing on the system. The biggest political barriers remaining are dealing with the influence the health insurance companies currently have on policy and figuring out how to pay for it. Estimates vary widely on how much this might cost but the range seems be \$1.4 trillion per year and \$2.8 trillion per year. Currently the United States spends \$1.4 on social health programs, so the additional cost to taxpayers would be between \$0 and \$1.4 trillion per year. Universal health care would have a wide-spread effect on Americans and there are more factors to consider than the number of youth living in poverty, but this should be an important part of the decision.

# Appendix

This is from the initial data cleanup.  This was done in many pieces to ensure the process was going as expected and to allow the other group members to follow the process step by step.

```{r, eval = FALSE}
#loading in the dataset
var_data <- read_csv("C:/Users/dpalc/OneDrive/Documents/School/DataWrangling/D2P_project/05b_analysis_file_update.csv")
dim(var_data)
```


```{r, eval = FALSE}
#selecting relevant columns
var_data <- select(var_data, year, cocnumber, dem_pop_pop_census, dem_soc_ed_bach_acs5yr, dem_soc_ed_hsgrad_acs5yr, dem_soc_ed_lesshs_acs5yr, econ_labor_medinc_acs5yr, econ_labor_unemp_rate_BLS, cpi_2017, dem_health_ins_acs5yr, dem_soc_white_census, dem_soc_black_census, dem_soc_native_census, dem_soc_asian_census, dem_soc_pacific_census, dem_soc_hispanic_census)
```


```{r, eval = FALSE}
#removing data from other years
var_data <- filter(var_data, year == "2017")
```


```{r, eval = FALSE}
#selecting columns without the year
var_data <- select(var_data, cocnumber, dem_pop_pop_census, dem_soc_ed_bach_acs5yr, dem_soc_ed_hsgrad_acs5yr, dem_soc_ed_lesshs_acs5yr, econ_labor_medinc_acs5yr, econ_labor_unemp_rate_BLS, cpi_2017, dem_health_ins_acs5yr, dem_soc_white_census, dem_soc_black_census, dem_soc_native_census, dem_soc_asian_census, dem_soc_pacific_census, dem_soc_hispanic_census)
```


```{r, eval = FALSE}
#renaming columns
var_data <- rename(var_data, total_pop = dem_pop_pop_census, educ_bach = dem_soc_ed_bach_acs5yr, educ_hs = dem_soc_ed_hsgrad_acs5yr, educ_lesshs = dem_soc_ed_lesshs_acs5yr, med_income = econ_labor_medinc_acs5yr, unemployment = econ_labor_unemp_rate_BLS, cpi = cpi_2017, health_insurance = dem_health_ins_acs5yr, ethinicity_white = dem_soc_white_census, ethnicity_black = dem_soc_black_census, ethnicity_native = dem_soc_native_census, ethnicity_asian = dem_soc_asian_census, ethnicity_pacific = dem_soc_pacific_census, ethnicity_hispanic = dem_soc_hispanic_census)
```

The data we have is separated by state and CoC number.  We need to strip the CoC number from the state abbreviation, then combine each state's data into a single row.

```{r, eval = FALSE}
#isolating data in proportions
prop_data <- select(var_data, cocnumber, total_pop, educ_bach, educ_hs, educ_lesshs, unemployment, health_insurance)
```

```{r, eval = FALSE}
#changing proportion columns into total number
prop_data <- mutate(prop_data, educ_bach = total_pop*educ_bach)
```

```{r, eval = FALSE}
#changing all proportion columns into total number
prop_data <- mutate(prop_data, educ_hs = total_pop*educ_hs, educ_lesshs = total_pop*educ_lesshs, unemployment = total_pop*(unemployment/100), health_insurance = total_pop*health_insurance)
```

```{r, eval = FALSE}
#separating state code from CoC number
prop_data <- separate(prop_data, col = cocnumber, into = c("State_abb", "Coc_num"), sep = "-")
```

```{r, eval = FALSE}
#removing CoC number from dataset
prop_data <- select(prop_data, !Coc_num)
```

```{r, eval = FALSE}
#turning tibble into data table
prop_table <- data.table(prop_data)
```

```{r, eval = FALSE}
#combining rows by sum
prop_table <- aggregate(. ~State_abb, data = prop_table, sum)
```

```{r, eval = FALSE}
#writing to a file because I needed help from Chris
write.csv(prop_data, "C:/Users/dpalc/OneDrive/Documents/School/DataWrangling/D2P_project/data_to_add.csv")
```

```{r, eval = FALSE}
#separating columns that need to be averaged
avg_data <- select(var_data, cocnumber, med_income, cpi)
```

```{r, eval = FALSE}
#stripping the CoC number from the state abbreviation
avg_data <- avg_data %>% separate(col = cocnumber, into = c("State_abb", "Coc_num"), sep = "-") %>% select(State_abb, med_income, cpi)
```

```{r, eval = FALSE}
#combining rows by mean
avg_data <- aggregate(. ~State_abb, data = avg_data, mean)
```

```{r, eval = FALSE}
#isolating remaining columns to combine by row
sum_data <- select(var_data, cocnumber, ethinicity_white, ethnicity_black, ethnicity_native, ethnicity_asian, ethnicity_pacific, ethnicity_hispanic)
```

```{r, eval = FALSE}
#stripping the CoC number
sum_data <- sum_data %>% separate(col = cocnumber, into = c("State_abb", "Coc_num"), sep = "-") %>% select(State_abb, ethinicity_white, ethnicity_black, ethnicity_native, ethnicity_asian, ethnicity_pacific, ethnicity_hispanic) 
```

```{r, eval = FALSE}
#combining rows by sum
sum_data <- aggregate(. ~State_abb, data = sum_data, sum)
```

```{r, eval = FALSE}
#reforming var_data tibble
var_data_2 <- merge(x = prop_table, y = sum_data, by = "State_abb")
```

```{r, eval = FALSE}
#reforming var_data tibble continued
var_data <- merge(x = var_data_2, y = avg_data, by = "State_abb")
```

```{r, eval = FALSE}
#loading in poverty data, note: this was a basic dataset so I made some adjustments in excel before loading into R
pov_data <- read_csv("C:/Users/dpalc/OneDrive/Documents/School/DataWrangling/D2P_project/poverty_by_age.csv")
```

```{r, eval = FALSE}
#selecting relevant columns
pov_data <- select(pov_data, !total_pop_est)
```

```{r, eval = FALSE}
#merging the tibbles
final_data <- merge(x = pov_data, y = var_data, by = "State_abb")
```

```{r, eval = FALSE}
#writing the file
write.csv(final_data, "C:/Users/dpalc/OneDrive/Documents/School/DataWrangling/D2P_project/D2P_poverty_dataset.csv")
```

**This is the end of the initial data cleanup**
